Adaptive Safe Strong Screening for Efficient Lasso-Type Problems Optimization over the Solution Path

### 2020-09-15 main.tex
Change name from 'batch' to 'adaptive'

\title{Efficient Feature Screening for Lasso-Type
Problems via Batched Safe-Strong Rules}
\author{Chuyi Wang\\Department of Statistics and Actuarial Sciences\\University of Iowa
  \and
  Patrick Breheny\\Department of Biostatistics\\University of Iowa}
\date{}

@@

Because the optimal value of $\lambda$ is not known in advance, typical practice is to solve for $\beta$ along a sequence of values of the tuning parameter $\lambda$, known as the solution path.  Pathwise algorithms can be efficiently implemented by using the solution at a previous $\lambda$ value as a ``warm start'' for the next value of $\lambda$.  In particular, the pathwise coordinate descent algorithm\citep{friedman2007pathwise}, which optimizes $\beta(\lambda)$ one element at a time with the remaining coordinates held fixed, has been shown to outperform other lasso algorithms such as LARS (least angle regression)\citep{efron2004least}, especially in high dimensional settings where $p$ is large. In sparse settings, it scales up very well with a computational cost of approximately $O(np)$. As a result, it is widely used to fit the lasso and other penalized regression models.

@@

Previous screening rules can also be divided into two categories, sequential rules and one-step rules, by how the screening is carried out along the path of $\lambda$ values. Sequential rules do heavy computation for each $\lambda$ values, so they are slower but relatively powerful at each $\lambda$ values. One-step rules only require heavy computation at the beginning and thus are faster but they cannot remain powerful along the path for long.

In this paper, we proposed an adaptive batched screening scheme that changes how screening is done along the path of $\lambda$ values and implement it into hybrid approaches to form adaptive batched safe-strong rules. We divided the set of $\lambda$ values into batches and screening within each batch is based on the same solution at one $\lambda$ value at the beginning of the batch. This screening scheme helps avoid disadvantages of both sequential rules and one-step rules. On one hand, basing on the same solution avoids many costly computation within a batch and on the other hand, referring to solution at a close $\lambda$ value provides more screening power. An adaptive choice of batches sizes are applied using the screening performance from previous $\lambda$ values to better balance these two effects. It will be shown that our adaptive rule can greatly reduce the computation time and outperform other methods under all scenarios. Similar to hybrid approaches, our method can also utilize any member of safe rules, giving it the potential for further speed up given new safe rules and the flexibility to extend to other lasso type models. For example we extended the method to work with logistic regression with lasso penalty.

@@

\begin{itemize}
    \item A new feature screening framework (Section ?3.2??), which does screening in batches of $\lambda$ values with any safe screening rules, which can generate a family of more efficient and scalable screening rule, compared to previous screening rules that are either sequential or one-step.
    \item Two specific applications of this framework, Batch-SSR-SEDPP and Batch-SSR-Slores, for fitting lasso-penalized linear regression and logistic regression models, respectively (Section ?3.3??).
    \item An algorithm to adaptively determine efficient batch sizes (Section \ref{sec:batch-size}).
    \item Simulation studies (Section~\ref{sec:sim}) and real data analysis (Section~\ref{sec:real-data}) under a variety of scenarios, including testing data sets larger than memory, which showed that the proposed algorithms outperform other approaches in all cases, with substantial speed up.
    \item The Batch-SSR-SEDPP and Batch-SSR-Slores were implemented and added as new options to the publicly accessible \textbf{R} package \textbf{biglasso}\citep{zeng2017biglasso}, which was used to carry out the analyses in this manuscript.
\end{itemize}

@@

Before introducing current screening rules, there is also a similar idea to reduce the number of features for part of computation, called active-set cycling (AC) \citep{lee2007efficient}. When solving the coefficients for at $\lambda_{k+1}$, first only consider features whose coefficients were non-zero at $\lambda_k$ and discard other features with zero coefficients at $\lambda_k$. We compute the solution for the smaller model with these reduced number of coefficients.

@@

BEDPP is much faster, but at each $\lambda_k$, it is bounding $r(\lambda_k)$ by comparing it to $r(\lambda_0)=y$. As $\lambda_k$ gets further away from $\lambda_0$, this bound becomes less precise and the screening rule becomes less powerful. SEDPP is slower, but at each $\lambda_{k+1}$, it is comparing $r(\lambda_{k+1})$ to $r(\lambda_k)$. Because $\lambda_{k+1}$ and $\lambda_k$ are close, the bound will be tight and screening will keep being powerful.

@@

\section{Batched safe-strong rule}
\label{sec:method}

In this section, we will introduce the batched safe-strong rules, combine them with pathwise coordinate descent algorithm and propose a strategy to adaptively start new batches.

\subsection{Batched EDPP rule}

From the SEDPP rule (10) in Theorem 2.2, we can do screening at $\lambda_{k+1}$ given the solution at $\lambda_k$. Now consider we are solving the lasso problem along the path $\lambda_0,...,\lambda_k,\lambda_{k+2},...,\lambda_K$ instead. Using the SEDPP rule, we can do screening at $\lambda_{k+2}$ as well, given the solution at $\lambda_k$, which is considered the previous solution now. Similarly, given the solution at $\lambda_k$, we can do screening for a batch of $B$ $\lambda$-values: $\lambda_{k+1},...,\lambda_{k+B}$ and then all the discarded features are guaranteed to be zero. Batched EDPP is still a safe rule, but it is neither sequential nor one-step.

Similar to SEDPP, the batched EDPP rule can be stated in the following theorem:

\begin{theorem}[Batched EDPP]
    For the lasso problem (1), let $\lambda_0:=\max_j|x_j^Ty/n|$. Suppose we have a sequence of $\lambda$ values $\lambda_0>\lambda_1>...>\lambda_K$. Under condition (2)
    \begin{enumerate}
        \item For any $k=1,2,...,K-1$, given $\hat{\beta}(\lambda_k)$, $r(\lambda_k)$ and $\hat{y}(\lambda_k):=X\hat{\beta}(\lambda_k)$, then for some $B$ and any $b$ s.t. $k<k+b\leq k+B\leq K$, $\hat{\beta}_j(\lambda_{k+b})=0$ if
        \begin{equation}
            \begin{split}
                &\left|2\lambda_{k+b}x_j^Tr(\lambda_k)+(\lambda_k-\lambda_{k+b})\left( x_j^Ty-\frac{y^T\hat{y}(\lambda_k)x_j^T\hat{y}(\lambda_k)}{||\hat{y}(\lambda_k)||_2^2}\right)\right|\\&<2n\lambda_k\lambda_{k+b}-(\lambda_k-\lambda_{k+b})\sqrt{n||y||_2^2-\frac{n(y^T\hat{y}(\lambda_k))^2}{||\hat{y}(\lambda_k)||_2^2}}
            \end{split}
        \end{equation}
        \item For k=0, i.e., $\lambda_k=\lambda_0$, then for some $B$ and any $b$ s.t. $0<b\leq B\leq K$, $\hat{\beta}_j(\lambda_{k+b})=0$ if
        \begin{equation}
        |(\lambda_0+\lambda_b)x_j^Ty-(\lambda_0-\lambda_b)sign(x_*^Ty)\lambda_0x_j^Tx_*|<2n\lambda_0\lambda_b-(\lambda_0-\lambda_b)\sqrt{n||y||_2^2-n^2\lambda_0^2}.
    \end{equation}
    \end{enumerate}
\end{theorem}

For each $\lambda$-value in the batch $\lambda_{k+1},...,\lambda_{k+B}$, the computation in (11) or (12) is the same, except for one quantity $\lambda_{k+b}$ ($\lambda_b$ in the case $k=0$). The most expensive computations are $x_j^Tr(\lambda_k)$, $x_j^Tx_*$ and $x_j^Ty$, for $j=1,2,...,p$. $x_j^Ty$ and $x_j^Tx_*$ are the same for all $\lambda$-values, $x_j^Tr(\lambda_k)$ are the same for all $\lambda$-values within the batch and $x_j^T\hat{y}(\lambda_k)=x_j^Ty-x_j^Tr(\lambda_k)$, so once they are computed, they can be stored and reused to avoid more computation in that batch. These two computations take $O(np)$ time and the rest computation in (11) and (12) take at most a lower order of $O(p)$ time.

Compared to sequential methods like SEDPP, the batched method only requires one $O(np)$ computation in one batch, while SEDPP requires one $O(np)$ computation for each $\lambda$-value. Compared to one-step methods like BEDPP, if the batch size is not too large, the solution at the batch head $\lambda_k$ and the solution at some $\lambda_{k+b}$ in the batch are usually relatively close, so screening at $\lambda_{k+b}$ using solution at $\lambda_k$ will remain powerful. For one-step method, as $\lambda_k$ gets further away from $\lambda_0$, the screening power can drop drastically and
may not be any beneficial after that.

\subsection{Batched strong rule after batched EDPP rule}

Because HSSR \citep{zeng2017efficient} have been shown to provide significant speedup compared to safe rules alone, we should also consider adding a strong rule screening after the batched EDPP rule screening. The batched SSR and batched EDPP rule combined will form our batched safe strong rule.

For ordinary SSR, the quantities $x_j^Tr(\lambda_k)$ used in (4) is already computed in the previous post-convergence check in (2). However, combined with safe rules, previous post-convergence check only needs to check features that are not discarded at the previous $\lambda$-value. Thus, for features that are discarded at the previous $\lambda$-value but not at the current $\lambda$-value, their corresponding $x_j^Tr(\lambda_k)$ have not been computed yet and thus cannot be used without additional cost. Similar to the idea of BASIL \citep{qian2019fast}, we can also do the strong rule screening in batches to reduce the number of required computations.

We are going to combine batched EDPP rule and batched SSR to form the batch safe strong rule as following. First, we determine a batch $\lambda_{k+1},...,\lambda_{k+B}$, and both batched EDPP rule and batched SSR will be conducted in this batch. Second $x_j^Tr(\lambda_k)$ will be computed and stored. Then for each $\lambda_{k+b}$ in the batch, after doing batched EDPP rule screening, a batched SSR screening is carried out on the features that have not been discarded by batched EDPP rule and discard features that satisfies: $|x_j^Tr(\lambda_k)/n|<2\lambda_{k+b}-\lambda_k$. Last, a post-convergence check is conducted on the features that were discarded by batched SSR but not discarded by batched EDPP rule.

By choosing the same batch for batched EDPP rule and batched SSR, the quantity $x_j^Tr(\lambda_k)$ only needs to be computed once and can be used by both rules. Also, because batched EDPP rule can stay powerful, it keeps the set of features that require post-convergence check small, while because basic EDPP loses power for $\lambda$-values that are far away from $\lambda_0$, post-convergence check for HSSR will be more expensive.

\subsection{Coordinate descent with batched safe strong rule}

In this section, we will show the algorithm to do pathwise coordinate descent with our batched safe strong rule screening. We will only show the algorithm for solving one batch $\lambda_{k+1},...,\lambda_{k+B}$. After finishing this batch, we just need to start a new batch and apply the same algorithm using the output from last batch as input. Note here for the first batch with batch head $\lambda_0$, the residuals vector $r(\lambda_0)$ is just the response $y$, because all the $\hat{\beta}(\lambda_0)$ are zero. $\{x_j^Ty\}_{j=1}^p$ need to be computed at the beginning and stored for the whole path.

\begin{algorithm}[H]
    \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Pre}{Pre-compute}\SetKwInOut{Initialize}{Initialize}
    \SetAlgoLined
    \Input{$\{x_j\}_{j=1}^p, \{x_j^Ty\}_{j=1}^p, \hat{\beta}(\lambda_k), r(\lambda_k), \lambda_{k+1},...,\lambda_{k+B}$}
    \Pre{ $\{x_j^Tr(\lambda_k)\}_{j=1}^p, \{x_j^T\hat{y}(\lambda_k)=x_j^Ty-x_j^Tr(\lambda_k)\}_{j=1}^p, \hat{y}(\lambda_k)=y-r(\lambda_k)$}
    \BlankLine
    \For{$b\xleftarrow{}$ 1 \KwTo B}{
        Safe Screening: $\mathcal{S}_{k+b}\xleftarrow{}\{j:x_j$ not discarded by batched EDPP rule$\}$.

        Strong Screening: $\mathcal{R}_{k+b}\xleftarrow{}\{j\in \mathcal{S}_{k+b}:x_j$ not discarded by batched SSR$\}$.

        \Initialize{$\hat{\beta}(\lambda_{k+b})=\hat{\beta}(\lambda_{k+b-1}), r(\lambda_{k+b})=r(\lambda_{k+b-1})$}

        \Repeat{Post-convergence check passes}{
            Pathwise coordinate descent in Algorithm 1 with only features in $\mathcal{R}_{k+b}$ and with warm start $\hat{\beta}(\lambda_{k+b}), r(\lambda_{k+b})$, until converge. Save the outputs in $\hat{\beta}(\lambda_{k+b}), r(\lambda_{k+b})$.

            Post-convergence check for all $j\in \mathcal{S}_{k+b}\setminus\mathcal{R}_{k+b}$,

            \uIf{Post-convergence check does not pass}{
                Add violating features to $\mathcal{R}_{k+b}$\;
            }
        }
    }
    \Output{$\{\hat{\beta}(\lambda_{k+b})\}_{b=1}^B, r(\lambda_{k+B})$}
    \caption{Pathwise coordinate descent with batched safe strong rule screening}
\end{algorithm}

\subsection{Adaptive batched safe strong rule}

In this section we will develop an algorithm to adaptively determining when to end the current batch and start a new batch, by evaluating screening power at previous $\lambda$-values in current batch. Our goal is to minimize the computation cost, so we will first analyse computation complexity for each part in Algorithm 2. Then we will propose an adaptive algorithm that can minimize the entire computation cost.

@@

This complexity analysis will only consider calculations with $O(np)$ complexity. The rest calculations, those with $O(p)$ complexity for example, are not only faster in order, but also require no out-of-core computing. Calculation of ${x_j^Ty}_{j=1}^p$ is needed for pathwise coordinate descent in algorithm 1 and is not additional cost for screening, so it would not be considered either. We will assume that strong rule screening is always powerful in the sense that $|\mathcal{R}_b|=o(p)\approx0$, but safe rule screening may not. We also assume that post-convergence check will mostly pass in the first iteration and thus is only needed approximately once per $\lambda$-value.

Under these assumptions, there are two dominating parts in the algorithm for each batch. First, the computation of $\{x_j^Tr(\lambda_k)\}_{j=1}^p$ at the beginning of the batch, which requires $np$ multiplications. Then, for each $\lambda$-value $\lambda_{k+b}$, the post-convergence check requires approximately $|\mathcal{S}_b\setminus\mathcal{R}_b|\approx|\mathcal{S}_b|$ KKT condition checks and each KKT condition check requires $n$ multiplications. These sum up to $n\sum_{b=1}^B|\mathcal{S}_b|$ multiplications. Combining these two parts, we will have the average cost per $\lambda$-value in the batch $\lambda_{k+1},...,\lambda_{k+B}$ to be

\begin{equation}
    \bar{T}_k(B) \approx n\frac{p}{B}+n\frac{\sum_{b=1}^B|\mathcal{S}_b|}{B}.
\end{equation}

The first term is the average cost for safe screening. Increasing batch size $B$ means reusing calculation at the batch head more times for the safe screening and thus reducing the average cost. The second term is the average cost for KKT condition checking. As the batch size $B$ increases, safe rule screening becomes less powerful and the number of KKT condition checks required ($|\mathcal{S}_b|$) increases. The following plot illustrates the effect of $B$ on $|\mathcal{S}_b|$ in a real data set solved over a path of $K=100$ $\lambda$-values. The y axis is the proportion of discarded features, calculated as $1-|\mathcal{S}_b|/p$.

\begin{figure}[H]
    \centering
    \includegraphics[scale = 0.5]{plots/batchsizes.png}    \caption{Proportion of features rejected by screening rules vs index of $\lambda$-values for different batch sizes.}
    \label{fig:3.4.1}
\end{figure}

BEDPP (Basic EDPP) rule is the special case of batched EDPP with only one batch of the largest possible size $B=K=100$. When the batch is large, the safe rule screening will eventually fail to discard any features and make the KKT condition checking cost increase to its maximum $np$. SEDPP rule is the special case where all batches have the smallest possible batch size $B=1$. It remains very powerful along the path, making the KKT condition checking cost almost 0, but the safe screening itself is costly and takes $np$ time per $\lambda$-value. Batched EDPP can strike a balance between BEDPP and SEDPP. In this example, using batches with fixed sizes $B=10$ can reduce the safe screening cost by 90\% compared to SEDPP and reduce the KKT checking cost by 50\% to more than 90\% compared BEDPP.

This example also shows the performance of batch EDPP rule changes along the path of $\lambda$-values. Take the batched EDPP with $B=10$ as an example. For the first batch of 10 $\lambda$-values, the batched EDPP stays powerful even at the end, which means this batch is still exploitable. We may want to extend this batch to reduce average safe screening cost without harming the KKT condition checking cost much. For the last batch however, the screening power is close to 0 for the last few $\lambda$-values, which means this batch becomes useless at this point. Starting a new batch earlier can greatly reduce the KKT condition checking cost with relatively small increase in safe screening cost. Also, the performance of batch EDPP rule also depends on many factors such as the data, the range of $\lambda$-values and spacing between $\lambda$-values. These suggest that starting a new batch after every fixed number of $\lambda$-values cannot not be consistently efficient and we want a method that can automatically determine whether to start a new batch given current performance.

\subsubsection{Adaptive batch starting algorithm}

From previous example, we assume the average cost is a convex function of batch size $B$ which decreases first and then after some optimal value increases again. If we want to minimize the average cost per $\lambda$-value, we need to end the current batch when extending the current batch by 1 $\lambda$-value will increase the average cost. That is, end the current batch at $\lambda_{B-1}$ when $\bar{T}_k(B)>\bar{T}_k(B-1)$. From (13), this condition will be:

>>>>

\title{Adaptive Safe Strong Screening for Efficient Lasso-Type
Problems Optimization over the Solution Path}
\author{Chuyi Wang\\Department of Statistics and Actuarial Sciences\\University of Iowa
  \and
  Patrick Breheny\\Department of Biostatistics\\University of Iowa}
\date{}

@@

Because the optimal value of $\lambda$ is not known in advance, typical practice is to solve for $\beta$ along a sequence of values of the tuning parameter $\lambda$, known as the solution path.  Pathwise algorithms can be efficiently implemented by using the solution at a previous $\lambda$ value as a ``warm start'' for the next value of $\lambda$.  In particular, the pathwise coordinate descent algorithm\citep{friedman2007pathwise}, which optimizes $\beta(\lambda)$ one element at a time with the remaining coordinates held fixed, has been shown to outperform other lasso algorithms such as LARS (least angle regression)\citep{efron2004least}, especially in high dimensional settings where $p$ is large. In sparse settings, it scales up very well with a computational cost of approximately $O(np)$.

@@

Previous safe screening rules can also be divided into two categories, sequential rules and one-step rules, by how the screening is carried out along the path of $\lambda$ values. Sequential rules do heavy computation for each $\lambda$ values using solution from its previous $\lambda$ value, so they are slower but relatively powerful at each $\lambda$ values. One-step rules only require heavy computation at the beginning and thus are faster but they cannot remain powerful along the path for long.

In this paper, we proposed an adaptive safe screening scheme that changes how screening is done along the path of $\lambda$ values. We do safe screening on $\lambda$ values alone the path referencing to the same solution at the $\lambda$ value in the beginning, as in one-step rules, until it is beneficial to update to a new $\lambda$ value as the reference. This screening scheme helps avoid disadvantages of both sequential rules and one-step rules. On one hand, reusing the same solution as reference avoids many costly computation before updating and on the other hand, referring to solution at a closer $\lambda$ value provides more screening power. An adaptive scheme determines when to update the reference based on the previous screening performance, to better balance these two effects. It will be shown that our adaptive rule can greatly reduce the computation time and outperform other methods under all scenarios. Following the hybrid approaches, we also combined our approach with strong rules to form adaptive safe strong rules. Similar to hybrid approaches, our method can also utilize any member of safe rules, giving it the potential for further speed up given new safe rules and the flexibility to extend to other lasso type models. For example we extended the method to work with logistic regression with lasso penalty and group lasso.

@@

\begin{itemize}
    \item A new feature screening framework (Section ?3.2??), which updates the reference for screening alone the grid of $\lambda$ values on any combination of strong screening rules and safe screening rules, which can generate a family of more efficient and scalable screening rules, compared to previous screening rules that are either sequential or one-step.
    \item Three specific applications of this framework, Ada-SSR-EDPP, Ada-SSR-Slores and Ada-SSR-EDPP for group lasso, for fitting lasso-penalized linear regression and logistic regression models, respectively (Section ?3.4,4.1??).
    \item An algorithm to determine when to update to a new reference $\lambda$ value for best efficiency (Section \ref{sec:adaptive}).
    \item Simulation studies (Section~\ref{sec:sim}) and real data analysis (Section~\ref{sec:real-data}) under a variety of scenarios, including testing data sets larger than memory, which showed that the proposed algorithms outperform other approaches in all cases, with substantial speed up.
    \item The Ada-SSR-SEDPP and Ada-SSR-Slores were implemented and added as new options to the publicly accessible \textbf{R} package \textbf{biglasso}\citep{zeng2017biglasso}, which was used to carry out the analyses in this manuscript.
\end{itemize}

@@

Before introducing current screening rules, there is also a similar idea to reduce the number of features for part of computation, called active-set cycling (AC) \citep{lee2007efficient}. It can be implemented on top of any lasso algorithm with no additional cost. When solving the coefficients for at $\lambda_{k+1}$, first only consider features whose coefficients were non-zero at $\lambda_k$ and discard other features with zero coefficients at $\lambda_k$. We compute the solution for the smaller model with these reduced number of coefficients.

@@

BEDPP is much faster, but at each $\lambda_k$, it is bounding $r(\lambda_k)$ using $r(\lambda_0)=y$ as a reference. As $\lambda_k$ gets further away from $\lambda_0$, this bound becomes less precise and the screening rule becomes less powerful. SEDPP is slower, but at each $\lambda_{k+1}$, it uses $r(\lambda_{k})$ as the reference. Because $\lambda_{k+1}$ and $\lambda_k$ are close, the bound will be tight and screening will keep being powerful.

@@

\section{Adaptive safe-strong rule}
\label{sec:method}

In this section, we will introduce the framework of adaptive safe-strong rules, implement them with EDPP as the safe rule to form Ada-SSR-EDPP and propose a strategy to adaptively updating the reference.

\subsection{Adaptive EDPP safe rule}

From the SEDPP rule (10) in Theorem 2.2, we can do screening at $\lambda_{k+1}$ given the solution at $\lambda_k$. Now consider we are solving the lasso problem along the path $\lambda_0,...,\lambda_k,\lambda_{k+2},...,\lambda_K$ instead. Using the SEDPP rule, we can do screening at $\lambda_{k+2}$ as well, given the solution at $\lambda_k$, which is considered the previous solution now. Similarly, with solution at $\lambda_k$ as the reference, we can keep screening for a number of $\lambda$-values: $\lambda_{k+1},\lambda_{k+2},...$ until it is beneficial to update the solution at some $\lambda_{K+B}$ as the new reference and in this way all the discarded features are guaranteed to be zero. Adaptive EDPP is still a safe rule, but it is neither sequential nor one-step.

Similar to SEDPP, the Adaptive EDPP rule can be stated in the following theorem:

\begin{theorem}[Adaptive EDPP]
    For the lasso problem (1), let $\lambda_0:=\max_j|x_j^Ty/n|$. Suppose we have a sequence of $\lambda$ values $\lambda_0>\lambda_1>...>\lambda_K$. Under condition (2)
    \begin{enumerate}
        \item For any $k=1,2,...,K-1$, given $\hat{\beta}(\lambda_k)$, $r(\lambda_k)$ and $\hat{y}(\lambda_k):=X\hat{\beta}(\lambda_k)$, then for some $B$ and any $b$ s.t. $k<k+b\leq k+B\leq K$, $\hat{\beta}_j(\lambda_{k+b})=0$ if
        \begin{equation}
            \begin{split}
                &\left|2\lambda_{k+b}x_j^Tr(\lambda_k)+(\lambda_k-\lambda_{k+b})\left( x_j^Ty-\frac{y^T\hat{y}(\lambda_k)x_j^T\hat{y}(\lambda_k)}{||\hat{y}(\lambda_k)||_2^2}\right)\right|\\&<2n\lambda_k\lambda_{k+b}-(\lambda_k-\lambda_{k+b})\sqrt{n||y||_2^2-\frac{n(y^T\hat{y}(\lambda_k))^2}{||\hat{y}(\lambda_k)||_2^2}}
            \end{split}
        \end{equation}
        \item For k=0, i.e., $\lambda_k=\lambda_0$, then for some $B$ and any $b$ s.t. $0<b\leq B\leq K$, $\hat{\beta}_j(\lambda_{k+b})=0$ if
        \begin{equation}
        |(\lambda_0+\lambda_b)x_j^Ty-(\lambda_0-\lambda_b)sign(x_*^Ty)\lambda_0x_j^Tx_*|<2n\lambda_0\lambda_b-(\lambda_0-\lambda_b)\sqrt{n||y||_2^2-n^2\lambda_0^2}.
    \end{equation}
    \end{enumerate}
\end{theorem}

For each $\lambda$-value in one update $\lambda_{k+1},...,\lambda_{k+B}$, the computation in (11) or (12) is the same, except for one quantity $\lambda_{k+b}$ ($\lambda_b$ in the case $k=0$). The most expensive computations are $x_j^Tr(\lambda_k)$, $x_j^Tx_*$ and $x_j^Ty$, for $j=1,2,...,p$. $x_j^Ty$ and $x_j^Tx_*$ are the same for all $\lambda$-values, $x_j^Tr(\lambda_k)$ are the same for all $\lambda$-values within the update and $x_j^T\hat{y}(\lambda_k)=x_j^Ty-x_j^Tr(\lambda_k)$, so once they are computed, they can be stored and reused to avoid more computation in that update. These two computations take $O(np)$ time and the rest computation in (11) and (12) take at most a lower order of $O(p)$ time.

Compared to one-step methods like BEDPP, after updating, the solution as the reference at $\lambda_k$ and the solution at some $\lambda_{k+b}$ in that update are usually relatively close, so screening at $\lambda_{k+b}$ using solution at $\lambda_k$ will remain powerful. For one-step method, as $\lambda_k$ gets further away from $\lambda_0$, the screening power can drop drastically and
may not be any beneficial after that. Compared to sequential methods like SEDPP, the adaptive method only requires one $O(np)$ computation in one update, while SEDPP requires one $O(np)$ computation for each $\lambda$-value. The framework of adaptive safe screening can also be easily applied to other sequential safe rules, to form a family of adaptive safe rules to achieve the same improvement.

\subsection{Adaptive safe strong rule on top of adaptive EDPP rule}

Because HSSR \citep{zeng2017efficient} have been shown to provide significant speedup compared to safe rules alone, we should also consider adding a strong rule screening after the adaptive EDPP rule screening. The SSR and adpative EDPP rule combined will form our adaptive safe strong rule: Ada-SSR-EDPP.

This combination is not trivial. For ordinary SSR, the quantities $x_j^Tr(\lambda_k)$ used in (4) is already computed in the previous post-convergence check in (2). However, combined with safe rules, previous post-convergence check only needs to check features that are not discarded at the previous $\lambda$-value. Thus, for features that are discarded at the previous $\lambda$-value but not at the current $\lambda$-value, their corresponding $x_j^Tr(\lambda_k)$ have not been computed yet and thus cannot be used without additional cost. Similar to the idea of BASIL \citep{qian2019fast}, we can also do the strong rule screening adaptively by using the latest $x_j^Tr(\lambda_k)$ available, which is computed by adaptive EDPP rule at the latest update, to reduce the number of required computations.

We are going to combine adaptive EDPP rule and SSR to form the Ada-SSR-EDPP as following. Suppose we have a latest updated reference $\lambda_k$ with $x_j^Tr(\lambda_k)$ computed and want the solution at some following $\lambda_{k+b}$. First, we do adaptive EDPP rule screening. Second, an adaptive SSR screening is carried out on the features that have not been discarded by adaptive EDPP rule and discard features that satisfies: $|x_j^Tr(\lambda_k)/n|<2\lambda_{k+b}-\lambda_k$. Third, a post-convergence check is conducted on the features that were discarded by adaptive SSR but not discarded by adaptive EDPP rule. Then we move on to the next $\lambda_{k+b+1}$ and repeat the same procedure, unless we determine an update is needed.

By choosing the same reference for adaptive EDPP rule and adaptive SSR, the quantity $x_j^Tr(\lambda_k)$ only needs to be computed once and can be used by both rules. Also, because adaptive EDPP rule can stay powerful, it keeps the set of features that require post-convergence check small, while because basic EDPP loses power for $\lambda$-values that are far away from $\lambda_0$, post-convergence check for HSSR will be more expensive.

\subsection{Pathwise lasso algorithm with adaptive safe strong rule}

In this section, we will show the algorithm to solve the lasso problem over the solution path with pathwise coordinate descent algorithm and our adaptive safe strong screening rule. The pathwise coordinate descent algorithm can be substituted by any other pathwise lasso algorithms and the adaptive EDPP can also be substituted by any other adaptive safe rules in our framework. Note here the first reference chosen will be $\lambda_0$, where the residuals vector $r(\lambda_0)$ is just the response $y$, because all the $\hat{\beta}(\lambda_0)$ are zero. $\{x_j^Ty\}_{j=1}^p$ need to be computed at the beginning and stored for the whole path.

\begin{algorithm}[H]
    \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Pre}{Pre-compute}\SetKwInOut{Initialize}{Initialize}
    \SetAlgoLined
    \Input{$\{x_j\}_{j=1}^p, \{x_j^Ty\}_{j=1}^p,  \lambda_{0},...,\lambda_{K}$}
    \BlankLine
    \Initialize{reference = $\lambda_0$}

    \For{$k\xleftarrow{}$ 1 \KwTo K}{
        Safe Screening: $\mathcal{S}_{k}\xleftarrow{}\{j:x_j$ not discarded by adaptive EDPP rule given the reference$\}$.

        Strong Screening: $\mathcal{R}_{k}\xleftarrow{}\{j\in \mathcal{S}_{k}:x_j$ not discarded by adaptive SSR given the reference$\}$.

        \Initialize{$\hat{\beta}(\lambda_{k})=\hat{\beta}(\lambda_{k-1}), r(\lambda_{k})=r(\lambda_{k-1})$}

        \Repeat{Post-convergence check passes}{
            Pathwise lasso algorithm with only features in $\mathcal{R}_{k}$ and with warm start $\hat{\beta}(\lambda_{k}), r(\lambda_{k})$, until converge.

            Save new $\hat{\beta}(\lambda_{k}), r(\lambda_{k})$.

            Post-convergence check for all $j\in \mathcal{S}_{k}\setminus\mathcal{R}_{k}$,

            \uIf{Post-convergence check does not pass}{
                Add violating features to $\mathcal{R}_{k}$\;
            }
        }
        \uIf{Update conditions met}{
            Reference = $\lambda_{k}$

            Compute and save: $\{x_j^Tr(\lambda_k)\}_{j=1}^p, \{x_j^T\hat{y}(\lambda_k)=x_j^Ty-x_j^Tr(\lambda_k)\}_{j=1}^p, \hat{y}(\lambda_k)=y-r(\lambda_k)$.}

    }
    \Output{$\{\hat{\beta}(\lambda_{k})\}_{k=1}^K, r(\lambda_{k+B})$}
    \caption{Pathwise lasso algorithm with adaptive safe strong rule screening}
\end{algorithm}

\subsection{Adaptive update algorithm}
\label{sec:adaptive}

In this section we will fill in line 11 in Algorithm 2. We will develop an algorithm to adaptively determining when update the reference for screening, by evaluating screening power at previous $\lambda$-values since the last update. Our goal is to minimize the computation cost, so we will first analyse computation complexity for each part in Algorithm 2. Then we will propose an adaptive algorithm that can minimize the entire computation cost.

@@

This complexity analysis will only consider calculations with $O(np)$ complexity. The rest calculations, those with $O(p)$ complexity for example, are not only faster in order, but also consume less memory and thus usually do no need out-of-core computing. Calculation of ${x_j^Ty}_{j=1}^p$ is needed only once for the whole solution path, so it would be neglected too. We will assume that strong rule screening is always powerful in the sense that $|\mathcal{R}_b|=o(p)\approx0$, but safe rule screening may not. We also assume that post-convergence check will mostly pass in the first iteration and thus is only needed approximately once per $\lambda$-value.

Under these assumptions, there are two dominating parts in the algorithm within each update. First, the computation of $\{x_j^Tr(\lambda_k)\}_{j=1}^p$ at the time of updating the reference, which requires $np$ multiplications. Then, for each $\lambda$-value $\lambda_{k+b}$, the post-convergence check requires approximately $|\mathcal{S}_{k+b}\setminus\mathcal{R}_{k+b}|\approx|\mathcal{S}_{k+b}|$ KKT condition checks and each KKT condition check requires $n$ multiplications. These sum up to $n\sum_{i=k}^{k+B}|\mathcal{S}_k|$ multiplications, where $k+B$ is the start of the next update. Combining these two parts, we will have the average cost per $\lambda$-value in the update $\lambda_{k+1},...,\lambda_{k+B}$ to be

\begin{equation}
    \bar{T}_k(B) \approx n\frac{p}{B}+n\frac{\sum_{b=1}^B|\mathcal{S}_{k+b}|}{B}.
\end{equation}

The first term is the average cost for safe screening. Delaying next update (increasing $B$) means reusing calculation at the previous update more times for the safe screening and thus reducing the average cost. The second term is the average cost for KKT condition checking. If the old reference is reused too many times, safe rule screening becomes less powerful and the number of KKT condition checks required ($|\mathcal{S}_{k+b}|$) increases.

To better illustrate this two effects, we considered the most naive and simple updating scheme: update the reference periodically after every other $B$ $\lambda$ values. The following plot illustrates the effect of update delay $B$ on $|\mathcal{S}_{k+b}|$ in a real data set solved over a path of $K=100$ $\lambda$-values. The y axis is the proportion of discarded features, calculated as $1-|\mathcal{S}_{k+b}|/p$.

\begin{figure}[H]
    \centering
    \includegraphics[scale = 0.5]{plots/batchsizes.png}    \caption{Proportion of features rejected by screening rules vs index of $\lambda$-values for different update delay.}
    \label{fig:3.4.1}
\end{figure}

BEDPP (Basic EDPP) rule is the special case of adaptive EDPP that never updates ($B=K$). When the reference does not get updated for a long time, the safe rule screening will eventually fail to discard any features and make the KKT condition checking cost increase to its maximum $np$. SEDPP rule is the special case that updates at every $\lambda$ value ($B=1$). It remains very powerful along the path, making the KKT condition checking cost small, but the safe screening itself is costly and takes $np$ time per $\lambda$-value. Adaptive EDPP can strike a balance between BEDPP and SEDPP. In this example, simply using fixed update period $B=10$ can reduce the safe screening cost by 90\% compared to SEDPP and reduce the KKT checking cost by 50\% to more than 90\% compared BEDPP.

This example also shows the timing of updates should be decided carefully and adaptively. For the first 10 $\lambda$-values, the safe rule stays powerful, which means the old reference is still exploitable. We may want delay the update to reduce average safe screening cost without harming the KKT condition checking cost much. For the last 20 $\lambda$-values however, the screening power drops to 0 quickly, which means the reference becomes useless quickly. Updating more frequently can greatly reduce the KKT condition checking cost with relatively small increase in safe screening cost. Also, the performance of adaptive EDPP rule also depends on many factors such as the data, the range of $\lambda$-values and spacing between $\lambda$-values. These suggest that simple updating rules like periodic updating cannot not be consistently efficient and we want a method that can automatically determine whether to update given current performance.

\subsubsection{Efficient updating algorithm}
